---
title: "Project-COVID-Analysis-Final"
author: "D. Garton"
date: "2023-08-15"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data Loading
<!-- Read in COVID Data -->
```{r get_cov_data}
##get current data in the four files
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <-
  c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")
library(tidyverse)
urls <- str_c(url_in,file_names)
urls
global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])
```
### Tidy Data
```{r tidying_data}
global_cases <- global_cases %>%
        pivot_longer(cols = 
                     -c('Province/State', 'Country/Region', Lat, Long),
                 names_to = "date",
                 values_to = "cases")

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State',
                        'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "deaths")

global_cases <- global_cases %>% 
     select(-c(Lat,Long))
global_deaths <- global_deaths %>%
     select(-c(Lat,Long))

#merge two files together by date&province&region
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date=mdy(date))

summary(global)


#remove cases that are =0
global <- global %>% filter(cases > 0)

#US Cases
US_cases

US_cases %>% 
     pivot_longer(cols = -(UID:Combined_Key),
                 names_to = "date",
                  values_to = "cases")

#tidy date and drop Lat/Long. Lat/Long may be useful for creating heat maps across the US.
US_cases <- US_cases %>% 
    pivot_longer(cols = -(UID:Combined_Key),
                 names_to = "date",
                 values_to = "cases") %>%
    select(Admin2:cases) %>%
    mutate(date = mdy(date)) %>%
    select(-c(Lat,Long_))

#Conduct the same transformations on US_deaths
US_deaths <- read_csv(urls[3])
US_deaths %>% 
     pivot_longer(cols = -(UID:Combined_Key),
                 names_to = "date",
                  values_to = "deaths")


US_deaths <- US_deaths %>% 
    pivot_longer(cols = -(UID:Population),
                 names_to = "date",
                 values_to = "deaths") %>%
    select(Admin2:deaths) %>%
    mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))

#Join US deaths and cases
US  <- US_cases %>%
    full_join(US_deaths)

summary(US)
summary(global)
```

### Analysis
``` {r analysis}
#add populations
#Combine, Province Stage Country_Region in Global Data
global_2 <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

#Need Population data for Global
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read.csv(uid_lookup_url) %>%
  select(-c(Lat,Long_, Combined_Key, code3, iso2, iso3, Admin2))

#remove NA from Population column to allow summarising
uid2 <- uid %>%
  mutate(Population = coalesce(Population, 0))

# Join with global_2
global_3 <- global_2 %>%
    left_join(uid, by=c("Province_State", "Country_Region")) %>%
    select(-c(UID,FIPS)) %>%
    select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

#join US data to global
US_data <- US %>% 
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths)) %>%
  ungroup()

Global_data <- global %>%
  group_by(Country_Region,date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths)) %>%
  ungroup()

all_data <- rbind(Global_data, US_data)
all_data

pop_data <- uid2 %>%
  group_by(Country_Region) %>%
  summarise(Population = sum(Population)) %>%
  ungroup()

#attach population data to all_data
final_df <- all_data %>%
  left_join(pop_data, by= "Country_Region")

#add in deaths per mill and cases per mill columns
final_df <- final_df %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  mutate(cases_per_mill = cases * 1000000 / Population)

#quick plot
final_df %>%
  filter(cases > 0 )%>%
  filter(Country_Region == "Afghanistan") %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 Afghanistan", y=NULL)

#new cases and new deaths
final_df <- final_df %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

#add year and month
final_df$year <- year(final_df$date)
final_df$month <- month(final_df$date)

#summarise by month and year
final_df_ym <- final_df %>%
  group_by(Country_Region, year, month) %>%
  summarise(new_cases = sum(new_cases), new_deaths = sum(new_deaths), ave_deaths = mean(deaths_per_mill), ave_cases = mean(cases_per_mill), Population = max(Population), cases = sum(cases), deaths = sum(deaths)) %>%
  ungroup()

#plot new deaths over time for China
final_df_ym %>% 
  filter(Country_Region == "China") %>%
  #filter(Country_Region == "United Kingdom") %>%
  filter(year == 2020) %>%
  ggplot(aes(x = month, y = new_deaths, fill = Country_Region)) +
  geom_bar(stat = "identity",
           position = "dodge")

#max data
df_totals <- final_df %>%
  group_by(Country_Region) %>%
  summarise(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_mill = max(cases_per_mill),
            deaths_per_mill = max(deaths_per_mill)) %>%
  filter(cases>0, population >0)
#plot deaths against cases
df_totals %>% ggplot()+
  geom_point(aes(x=cases, y=deaths))

```

### Modelling
```{r modelling}
#creating a model of countries
df_totals %>%
    slice_max(deaths_per_mill, n=10) %>%
    select(Country_Region,cases,cases_per_mill,deaths,deaths_per_mill,population)

model1 <- lm(deaths_per_mill ~ cases_per_mill, data = df_totals)
summary(model1)
df2 <- df_totals %>%
     mutate(pred = predict(model1))

df2 %>% ggplot() +
  geom_point(aes(x = cases_per_mill, y = deaths_per_mill), color = "blue") +
  geom_point(aes(x = cases_per_mill, y= pred), color = "red")
```

### Conclusion
Modelling the deaths_per_mill as a function of the number of cases per million does not create a strong model, this is most likely due to there being other factors influencing the number of deaths. The countries with the highest number of deaths per million of population seem to be mostly Eastern European which may be explained by a low financial status of these countries.

There are similar patterns between all countries for the covid 19 cases and deaths where there was a great initial spike around the start of 2020 which then slowly started to decline in rate over the following years/days. 
### Bias
I haven chosen to represent random countries throughout my analysis to look into the data more. However, the data being used may be biased in the fact that it could be incorrectly reported by certain countries to make themselves seem 'better' than they are. 